\Large{Research Seminar Software Science} \\ \huge{Simulation of monadic effects in Haskell}

### 2018-04-18 paper.tex
added monad laws definition

\documentclass[a4paper, twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage{blindtext}
\usepackage{cite}
\usepackage{comment}
\usepackage[a4paper,left=3cm,right=3cm,top=4cm,bottom=4cm]{geometry}

@@

A straightforward consequence of laziness is the evaluation order of a Haskell program. Being demand-driven, it becomes impossible for a function call to produce any side effects. Therefore, Haskell is considered to be a \textit{pure language}. For example, a function of type $Char \rightarrow Char$ will neither write nor read any mutable variables. Nevertheless, side effects are without any doubt very convenient. Haskell's restricted side effects resulted in a sluggish input/output. For this reason, monadic I/O was created; thus by leading to a functional language with side effects. 
\section{Monads}

There exists a certain complexity when it comes to grasping to concept of a monad. From history, monads were initially adopted in Haskell to produce input and output operations. However, the capabilities of monads were not bound to just I/O. Additionally, monads can support multiple operations such as state, non-determinism, continuations, exceptions etc. \\ \\ 
The default monad is composed by three esential components: 
\begin{itemize}
    \item a type constructor
    \item a function return
    \item a bind operator ($\gg$=)
\end{itemize}
The \textit{return} function and the bind operator ($\gg$=) have the following types:

\begin{center}
   return   :: a $\rightarrow$ m a \\
   ($\gg$=) :: m a $\rightarrow$ (a $\rightarrow$ m b) $\rightarrow$ m b 
   \end{center}
In order for the above function definitions to hold, the three monadic laws need to be satisfied.

The Maybe monad has been selected to provide a concrete insight on how a monad functions.

>>>>

\documentclass[a4paper, onecolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage{blindtext}
\usepackage{cite}
\usepackage{comment}
\usepackage{color}
\usepackage[a4paper,left=3cm,right=3cm,top=4cm,bottom=4cm]{geometry}

@@

A straightforward consequence of laziness is the evaluation order of a Haskell program. Being demand-driven, it becomes impossible for a function call to produce any side effects. Therefore, Haskell is considered to be a \textit{pure language}. For example, a function of type $Char \rightarrow Char$ will neither write nor read any mutable variables. Nevertheless, side effects are without any doubt very convenient. Haskell's restricted side effects resulted in a sluggish input/output. For this reason, monadic I/O was created; thus by leading to a functional language with side effects. 
\section{Monads}
Monads originate from a section of mathematics called category theory. Luckily for us, it is not required to have prior knowledge in category theory in order to comprehend the application of monads in Haskell. 
Nevertheless, there exists a certain complexity when it comes to grasping to concept of a monad. From history, monads were initially adopted in Haskell to produce input and output operations. However, the capabilities of monads were not bound to just I/O. Additionally, monads can support multiple operations such as state, non-determinism, continuations, exceptions etc. \\ \\ 
The default monad is composed by three essential components: 
\begin{itemize}
    \item a type constructor
    \item a function return
    \item a bind operator ($\gg$=)
\end{itemize}
The \textit{return} function and the bind operator ($\gg$=) have the following types:

\begin{center}
   return   :: a $\rightarrow$ m a \\
   ($\gg$=) :: m a $\rightarrow$ (a $\rightarrow$ m b) $\rightarrow$ m b 
   \end{center}
In order for the above function definitions to hold, the monad laws need to be satisfied. There exist three monadic laws:

\begin{itemize}
    \item right unit \hspace{20mm} m $\gg$= return \textcolor{red}{=} m 
    \item left unit \hspace{20mm} return x $\gg$= f \textcolor{red}{=} f x   
    \item associative \hspace{15mm} (m $\gg$= f) $\gg$= g \textcolor{red}{=} m $\gg$= (\textbackslash x $\rightarrow$ f x $\gg$= g) 
\end{itemize}
According to the laws of right and left unit, the return serves as a collector of a value where no computation occurs. As the type of \textit{return} suggests, it takes a value of type \textit{a}, puts it in a special context \textit{m} and returns as a result \textit{m a}.
The Maybe monad has been selected to provide a concrete insight on how a monad functions.
